\tit

### 2021-02-27 paper.tex
non-idempotence

%% \acmtrue

\ifacm
\documentclass[acmsmall,screen,anonymous,timestamp]{acmart}
\else
\documentclass[hidelinks]{article}  % fleqn,12pt
\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
\usepackage[useregional]{datetime2}
\fi

@@

\ifacm
\bibliographystyle{ACM-Reference-Format}
\else
\usepackage[round]{natbib} % [square]
\bibliographystyle{plainnat}
\fi

@@

\figrefdef{predicates}{Predicate operations}{
\ifacm
\small
\fi
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\ifacm
\vspace{1.5ex}
\else
\vspace{-2ex}
\fi
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set⇂} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} the type \AF{Set⇂}.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\ifacm
\vspace{1ex}
\else
\vspace{-2ex}
\fi
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).

@@

Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.
(The equivalence relation used for types is isomorphism rather than equality.)
Both are also closed.
For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{starˡ}.

@@

Of the previous work involving language derivatives, none appear to have based their formal specification and proofs on the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}, although \citet{Abel2016} did indeed mention this definition informally as motivation for the more complex, coinductively defined operations on tries.
\citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.

>>>>

\acmtrue

\ifacm
\documentclass[acmsmall,screen,anonymous,timestamp,review]{acmart}
\else
\documentclass[hidelinks]{article}  % fleqn,12pt
\usepackage[margin=1in]{geometry}  % 0.12in, 0.9in, 1in
\usepackage[useregional]{datetime2}
\fi

@@

\ifacm
\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}
\else
\usepackage[round]{natbib} % [square]
\bibliographystyle{plainnat}
\fi

@@

\figrefdef{predicates}{Predicate operations}{
\ifacm
\small
\fi
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\ifacm
\vspace{1.5ex}
\else
\vspace{-2ex}
\fi
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from \AF{Set ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\ifacm
\vspace{1ex}
\else
\vspace{-2ex}
\fi
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
%% (any domain type)
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
%% (any monoid)
\agda{domain-ops}
%% (lists)
\agda{list-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).

@@

Conveniently, booleans and types form commutative semirings with all necessary proofs already in the standard library.\footnote{The equivalence relation used for types is isomorphism rather than equality.
The boolean semiring is idempotent, while the type semiring is not (in order to distinguish multiple parsings in ambiguous languages).
The latter non-idempotence accounts for the difference between the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas} and the corresponding rules in previous work, as mentioned in \secref{Related Work}.}
They are both closed as well.
For booleans, closure maps both \AIC{false} and \AIC{true} to \AIC{true}, with the \AF{starˡ} law holding definitionally.
For types, the closure of \AB{A} is {\AB{A} ✶} (the usual inductive list type) with a simple, non-inductive proof of \AF{starˡ}.

@@

Of the previous work involving language derivatives, none appear to have based their formal specification and proofs on the essential, non-inductive, nearly-trivial definition in terms of functions of lists from \secref{Decomposing Languages}, although \citet{Abel2016} did indeed mention this definition informally as motivation for the more complex, coinductively defined operations on tries.
\citet{Brzozowski64} also made clear mention of the underlying meaning on sets of strings in his original work on regular expression derivatives.
Moreover, none of these previous investigations appear to have addressed proof isomorphism (distinguishing multiple parsings/explanations), and correspondingly all use transformations based the less precise laws originally discovered by \citet{Brzozowski64}.
Those laws were based on the assumption that union (``addition'' in both language semirings of \secref{Predicate Algebra}) is idempotent, which is not true for type-level language predicates related by proof isomorphism (rather than mere logical equivalence).
The lack of idempotence is crucial for going beyond recognizers to parsers, in which we might want to extract more than one proof (parsing).
The difference is mainly visible in the \AF ν and \AF δ lemmas for {\AB P ⋆ \AB Q} and {\AB P ☆} in \figref{nu-delta-lemmas}.
